An Improved Method for Coupling Hydrodynamics with Astrophysical Reaction Networks

### 2022-05-09 simplified_sdc_intro/paper.tex
lots of updates

A particularly difficult phase of evolution to model is the nuclear
statistical equilibrium that sets in for temperatures in excess of
$\mbox{few} \times 10^9~\mathrm{K}$.  Physically, the forward and
reverse rates of reaction should balance leading to an equilibrium.
With operator splitting, an NSE region will have a large positive flow
through the network in a zone in one step followed by a large negative
flow over the next timestep, as the code struggles to produce an
equilibrium.  These large changes in abundances (and large alternately
positive and negative energy generation rates) can be a challenge for
a code.  The easiest way to improve the coupling is to cut the
timestep (see, e.g., \citealt{couch:2015}), but this makes simulations
prohibitively expensive.  Sometimes the burning is simply halted on a
zone-by-zone basis when NSE conditions are reached (e.g., as in
\citealt{hedet}).  Alternately, at high temperatures, a reaction
network can be replaced with a table of NSE abundances and the zone's
composition set through table look-ups (e.g.\ \citealt{ma:2013}).  The
present work focuses on networks alone, but in a follow-on paper we
will explore hybrid burning models consisting of networks and NSE
tables.

@@

The time-integration approach presented here is considerably simpler
than the SDC method of \citet{castro_sdc}, but allows us to reuse the
main CTU \MarginPar{any cites to CCSE papers?}  hydrodynamics
construction and a largely similar ODE integration scheme, making this
method easier to add to existing simulation codes.  Furthermore, it
also extends to adaptive mesh refinement with subcycling in a
straightforward manner, avoiding the complications described in
\citep{mccorquodalecolella} needed to fill ghost cells when using
method-of-lines integration.  However, it is restricted to
second-order accuracy overall.  We term this algorithm the
``simplified-SDC method''.  In this paper we describe the overall
method and demonstrate it on a test problem and astrophysical
simulation using \castro.  All of the code to reproduce the results in
this paper are in the \castro\ github
repository\footnote{\url{https://github.com/amrex-astro/Castro/}}.

@@

We solve the Euler equations for compressible, reacting flow.  For ease
of exposition we describe the one-dimensional case;
multidimensional extensions are a straightforward modification to
include the CTU hydrodynamics scheme.  Our conserved variables are
\begin{equation}
  \Uc = \left ( \begin{array}{c}
           \rho \\
           \rho X_k \\
           \rho u \\
           \rho E \\
           \rho e \end{array}\right )
\end{equation}
where $\rho$ is the mass density, $u$ is velocity, $E$ is specific
total energy, $p$ is the pressure, and we carry nuclear species mass
fractions, $X_k$.  The specific total
energy relates to the specific internal energy, $e$, as $E = e + u^2/2$,
and we also separately evolve $e$, as part of a dual
energy formulation (see \citealt{bryan:1995,wdmergerI}).
 The mass fractions are constrained to sum to 1, $\sum_k X_k = 1$, but
 no such constraint exists on the auxiliary variables. 
Defining the hydrodynamical fluxes:
\begin{equation}
  \Fb(\Uc) = \left ( \begin{array}{c}
         \rho u \\
         \rho X_k u \\
         \rho u^2 p \\
         (\rho E + p) u \\
         \rho e u \end{array}\right )
\end{equation}
We can write the system in conservative form for all state variables aside from $(\rho e)$ as:
\begin{equation}
  \ddt{\Uc} + \ddx{\Fb(\Uc)} = \Sc(\Uc)
\end{equation}
where for the special case of $(\rho e)$, we have an additional ``pdV'' term:
\begin{equation}
\ddt{(\rho e)} + \ddx{F(\rho e)} + p \ddx{u} = S(\rho e)
\end{equation}

@@

To close the system, we need the equation of state.  For a system
where the composition is completely specified by the mass fractions,
$X_k$, the equation of state would take the form:
\begin{equation}
p = p(\rho, X_k, e)
\end{equation}

@@

The CTU+PPM method for hydrodynamics is second-order accurate in space
and time.  We want to couple the reactive sources to the hydrodynamics
to be second-order in time as well.  As discussed above, nuclear
reaction sources are stiff, and need to be integrated using implicit
methods for stabilty.  Operator splitting (e.g., Strang) is
traditionally employed here, and is used as a benchmark for comparison
in this paper.  We'll discuss this traditional approach next before moving
on to our new time-coupling method.  Figures~\ref{fig:flowchart_I} and \ref{fig:flowchart_II} show a graphical
flowchart of the reaction and hydrodynamics advances in the two methods.

@@

In Strang splitting, we first integrate the system with reactions terms only (no advection)
over $\Delta t/2$, then integrate the advection terms only (no reactions) over $\Delta t$,
and finally integrate the reaction terms only over $\Delta t/2$.

@@

Since this method is based off of the CTU hydrodynamics scheme, it
benefits from the larger timestep that method can take (when done with
full corner coupling, the advective CFL condition is unity) as
compared to a method-of-lines approach (see \citealt{ppmunsplit}).  In
addition to the standard CFL timestep limiter for explicit
hydrodynamics, timestep limiters based on
the energy generation or abundance changes over a timestep, such as
those introduced in \cite{prometheus}, are often used.  An energy limiter takes the
form of either:
\begin{equation}
\label{eq:dt:nuce_I}
\Delta t \le f_e\, \min_{i} \left\{\frac{e^n_{i}}{|\dot{S}_{i}^n|}\right\} \\
\end{equation}
or
\begin{equation}
\label{eq:dt:nuce_II}
\Delta t \le f_e\, \min_{i} \left\{ \frac{e^n_i}{|\int \dot{S}_i(t) dt|} \right\}
\end{equation}
where $i$ is the zone index and $f_e$ is a parameter used to control
the allowed change (e.g., $f_e = 0.5$ will limit the step such that
reactions can only increase the internal energy by 50\%).  The two
formulations differ in that Eq.~\ref{eq:dt:nuce_I} uses the
instantaneous evaluation of the energy generation rate at the start of
the timestep while Eq.~\ref{eq:dt:nuce_II} uses the integral of the
energy generation rate over the entire step (or last Strang half).

\castro\ has the ability to reject a timestep if it detects a failure
and retry with smaller timesteps (subcycling to make up the original
required timestep).  Among the conditions that can trigger this are
density falling below zero during advection, the ODE integration
failing to converge in the implicit solve (due to too many steps or
the internal timestep falling below a minimum), or violation of one of
the timestep limiters during the step.  This means that equation
(\ref{eq:dt:nuce_I}) or (\ref{eq:dt:nuce_II}) are not reactive, but
instead guaranteed to be met throughout the simulation, because the
step is rejected if they are violated.  This contrasts to similar
approaches used in other codes where conditions \ref{eq:dt:nuce_I} or
\ref{eq:dt:nuce_II} are used to restrict the next timestep size, but
the update over the current step is kept, which already violated the
constraints. \MarginPar{would be nice to have a reference here} The
retry mechanism in \castro\ works with both the Strang and
simplified-SDC integration scheme.  Retrying the step is much more
stringent that simply adjusting the next step, so we put a cap on the
number of retries allowed, aborting if we exceed the cap.

@@

Our simplified-SDC approach is based on the method described in \citet{SDC-old} and
a similar (but unpublished) implementation in the \maestroex\ simulation code \cite{maestroex}.
We again start by considering the update
in integral form:
\begin{equation}
\Uc^{n+1} = \Uc^n + \int_t^{t+\Delta t} \left [ \Advs{\Uc} + \Rb(\Uc) \right ] dt,
\end{equation}
In this approach, we 
approximate the advective term
piecewise constant in time, using the value at the midpoint in time,
to achieve second-order accuracy, giving us:
\begin{equation}
\label{eq:integral:simplesdc}
\Uc^{n+1} = \Uc^n + \int_t^{t+\Delta t} \left \{ \Adv{\Uc}^{n+1/2} + \Rb(\Uc) \right \} dt
\end{equation}
The \castro\ version of the simplified-SDC algorithm
differs from the previous versions due to the need to do some operations
on the conserved variable state and some on the primitive variable
state.

@@

We use a modified version of the VODE~\citep{vode} integrator,
designed for stiff ODEs.  Our version has been ported C++, as a
header-only library enabling it to run on GPUs.  We have also added
checks to the timestep rejection logic that help prevent VODE from
wandering too far off the solution\footnote{This modified version of
VODE is available in our Microphysics repo:
\url{https://github.com/starkiller-astro/Microphysics}.}  These have
been shown empirically to improve the performance of VODE.

@@

For Strang, we require that
an individual mass fraction not change too much over a step:
\begin{equation}
\label{eq:vode_species_reject}
\begin{array}{c}|X_k^{m+1}| < f |X_k^m| \\ |X_k^{m+1}| > \frac{1}{f} |X_k^m| \end{array} ~  \mbox{if}~ |X_k^m| > \eta \atolspec ~\mbox{and}~ |X_k^{m+1}| > \eta \atolspec 
\end{equation}
Here, $m$ is the current VODE solution and $m+1$ is the potential
new-time solution.  The parameter $f$ is the factor a mass fraction is
allowed to change per VODE step.  We use $f = 4$ for the results here.
The parameter $\eta$ allows us to only use this condition for species
with mass fractions above a threshold.  We use $\eta = 1$ by default,
but $\eta = f$ sometimes helps prevent VODE from worrying about
species that dip below the threshold during integration, since VODE
uses a norm over all of the integrated variables to compute the total
error.  If these conditions are violated, then VODE will reject the
timestep and retry with a smaller step.  We also require that the mass
fractions are contained in $[0, 1]$ to a tolerance of $10^{-2}$.

For SDC, we enforce these same constraints, but for
Eq.~\ref{eq:vode_species_reject}, we only use the change from $m$ to
$m+1$ due to reactions, by subtracting off the advective contribution
over that substep.  We also require that:
\begin{align}
\rho^{m+1} &> 0 \\
(\rho e)^{m+1} &> 0
\end{align}

For both Strang and simplified-SDC, we will use a numerical Jacobian.
VODE computes it internally for Strang and we use the method from
Appendix \ref{app:jacobian} for the simplified-SDC method.  VODE
employs Jacobian-caching, so the Jacobian does not need to be
re-evaluated each time it is used.

@@

\cite{castro_sdc} introduced a test problem for measuring convergence
of a reacting hydrodynamic algorithm.  This is based on the acoustic
pulse problem from \citep{mccorquodalecolella}, and was also used in
\citep{strang_rnaas}.  A periodic domain with a uniform entropy is
initialized with a pressure profile with a small perturbation.  This
drives a low amplitude acoustic wave radially outward from the center.
A simple reaction network with the triple-alpha and
$\isotm{C}{12}(\alpha,\gamma)\isotm{O}{16}$ rates releases energy.  By
running at different resolutions, with a fixed timestep that scales
with resolution, we can compute the convergence rate.  In order to
demonstrate second-order convergence, the problem must be smooth (so
the slope limiters don't kick in severely), which means that the
energy release cannot be so vigorous as to drive shocks.  test here
with the simplified SDC algorithm.

@@

Tables \ref{table:react_converge_strang} and
\ref{table:react_converge_sdc} show the Strang and simplified-SDC
convergence respectively.  We see that both methods converge nearly
second order, and agree well with each other.

@@

We next model a double detonation Type Ia
supernova.  In this model, a detonation begins in the accreted helium
layer on a sub-Chandrasekhar mass white dwarf and that detonation can
either drive a compression wave into the core igniting a second
detonation in the carbon / oxygen core, or burn through the interface
at the base of the accreted layer producing an inward propagating
carbon detonation.  Our goal here is not to explore the feasibility of
the model or understand whether a detonation propagating through the
He-C interface is physical, but rather to look at the coupling of the
hydrodynamics and reactions.  For that reason, we explore only a
single model and start off with a rather large temperature
perturbation in the He layer.

We use a network with 28 nuclei and 96 rates that captures He burning \MarginPar{should I do a few aprox13 runs for comparison?}
with links for $(\alpha,p)(p,\gamma)$ reactions and the rates
involving \isot{N}{14} to bypass
$\isotm{C}{12}(\alpha,\gamma)\isotm{O}{16}$ as described in
\cite{shenbildsten}.  The rates are taken from the ReacLib library
\citep{reaclib} and the network is written by
\pynucastro~\citep{pynucastro} directly in the C++ format that our
code requires.  The nuclei and links in the network are shown in
Figure~\ref{fig:subch_network}.  This moderate-sized network with many
reaction pathways (forward and reverse) make it a good test for the
coupling between hydro and reactions.  An overview of the network is
shown in Figure~\ref{fig:subch_network}.  This network is available 
as {\tt subch2} in \citep{microphysics}.

@@

An initial perturbation is placed on the symmetry axis of the form:

We do a suite of runs with both Strang and the simplified-SDC
integration, using CFL numbers of 0.2, 0.4, and 0.8 and reaction
network tolerances of $\atolspec = \rtolspec = 10^{-5}$ and $\atolener
= \rtolener = 10^{-6}$.  We also run two sets with tighter tolerances.
The runs labeled ``mixed\_tols'' have $\atolspec = 10^{-8}$ and the
runs labeled ``tol1.e-6'' have $\atolspec = \rtolspec = 10^{-6}$, with
the other tolerances unchanged.  For the step rejection part of VODE,
we use $f = 4$ and $\eta = 4$.

@@

\begin{deluxetable}{llll}
\tablecaption{\label{table:masses} Species masses.}
\tablehead{\colhead{simulation} & \colhead{$M_{\isotm{He}{4}}$} & \colhead{$M_{\isotm{Si}{28}}$} & \colhead{$M_{\isotm{Ni}{56}}$}}
\startdata
\multicolumn{4}{c}{Strang runs} \\
\hline
{\tt strang\_subch2\_cfl0.1}           & 1.059\times 10^{32}  &  3.804\times 10^{31}  &  1.092\times 10^{32} \\
{\tt strang\_subch2\_cfl0.2}           & 9.114\times 10^{31}  &  4.174\times 10^{31}  &  1.078\times 10^{32} \\
{\tt strang\_subch2\_cfl0.2\_tol1.e-8} & 7.222\times 10^{31}  &  8.623\times 10^{31}  &  4.041\times 10^{31} \\
{\tt strang\_subch2\_cfl0.4}           & 7.819\times 10^{31}  &  4.208\times 10^{31}  &  1.045\times 10^{32} \\
\hline
\multicolumn{4}{c}{SDC runs} \\
\hline
{\tt sdc\_subch2\_cfl0.1}           & 7.319\times 10^{31}  &  9.147\times 10^{31}  &  3.765\times 10^{31} \\
{\tt sdc\_subch2\_cfl0.2}           & 7.208\times 10^{31}  &  9.151\times 10^{31}  &  3.783\times 10^{31} \\
{\tt sdc\_subch2\_cfl0.2\_tol1.e-8} & 7.376\times 10^{31}  &  9.220\times 10^{31}  &  3.513\times 10^{31} \\
{\tt sdc\_subch2\_cfl0.4}           & 7.203\times 10^{31}  &  8.843\times 10^{31}  &  3.754\times 10^{31} \\
\hline
\enddata
\end{deluxetable}

1-d slices along x=0?

@@

\section{Reaction Network}

We build a reaction network that approximates an $\alpha$-chain,
including all $\isotm{He}{4}$, and all of the $\alpha$ nuclei from
$\isotm{C}{12}$ to $\isotm{Ni}{56}$.  We want to capture both the
$(\alpha,\gamma)$ and $(\alpha,p)(p,\gamma)$ links, so we include the
intermediate nuclei from the $(\alpha,p)$ rate instead of
approximating these links assuming proton equilibriation.  Following
\citet{shenbildsten}, we include the sequence
$\isotm{N}{14}(\alpha,\gamma)\isotm{F}{18}(\alpha, p)\isotm{Ne}{21}$,
which produce the protons that can allow for $\isotm{C}{12}(p,
\gamma)\isotm{N}{13}(\alpha, p)\isotm{O}{16}$ to compete with
$\isotm{C}{12}(\alpha,\gamma)\isotm{O}{16}$.  We add $\isotm{Na}{22}$
to link the $\isotm{Ne}{21}$ produced back to the $\alpha$-chain.  We
grab all of the ReacLib \citep{reaclib} reaction rates linking these
nuclei using \pynucastro~\citep{pynucastro}.  We are missing one rate
each for $\isotm{C}{12} + \isotm{C}{12}$, $\isotm{C}{12} +
\isotm{O}{16}$, and $\isotm{O}{16} + \isotm{O}{16}$.  These are
the sequences $\isotm{C}{12}(\isotm{C}{12}, \mathrm{n})\isotm{Mg}{23}(n, \gamma)\isotm{Mg}{24}$,
$\isotm{O}{16}(\isotm{C}{12}, \mathrm{n})\isotm{Si}{27}(n, \gamma)\isotm{Si}{28}$, and
$\isotm{O}{16}(\isotm{O}{16}, \mathrm{n})\isotm{S}{31}(n, \gamma)\isotm{S}{32}$,involving neutron
production followed by a capture back to one of our $\alpha$-chain
nuclei.  These are the only rates we approximate here, assuming
that the subsequent neutron capture is instantaneous.

>>>>

A particularly difficult phase of evolution to model is the nuclear
statistical equilibrium that sets in for temperatures in excess of
$\mbox{few} \times 10^9~\mathrm{K}$.  Physically, the forward and
reverse rates of reaction should balance leading to an equilibrium.
With operator splitting, an NSE region will have a large positive flow
through the network in a zone in one step followed by a large negative
flow over the next timestep, as the code struggles to produce an
equilibrium.  These large changes in abundances (and large alternately
positive and negative energy generation rates) can be a challenge for
a code.  The easiest way to improve the coupling is to cut the
timestep (see, e.g., \citealt{couch:2015}), but this can make simulations
prohibitively expensive.  Sometimes the burning is simply halted on a
zone-by-zone basis when NSE conditions are reached (e.g., as in
\citealt{hedet}).  The
present work focuses on networks alone, but in a follow-on paper we
will explore hybrid burning models consisting of networks and tables for
nuclear statistical equilibrium.

@@

The time-integration approach presented here is considerably simpler
than the SDC method of \citet{castro_sdc}, but allows us to reuse the
main CTU \MarginPar{any cites to CCSE papers?}  hydrodynamics
construction and a largely similar ODE integration scheme, making this
method easier to add to existing simulation codes.  Furthermore, it
also extends to adaptive mesh refinement with subcycling in a
straightforward manner, avoiding the complications described in
\citep{mccorquodalecolella} needed to fill ghost cells when using
method-of-lines integration.  However, it is restricted to
second-order accuracy in time overall.  We term this algorithm the
``simplified-SDC method''.  In this paper we describe the overall
method and demonstrate it on a test problem and astrophysical
simulation using \castro.  All of the code to reproduce the results in
this paper are in the \castro\ github
repository\footnote{\url{https://github.com/amrex-astro/Castro/}}.

@@

We solve the Euler equations for compressible, reacting flow.  For ease
of exposition we describe the one-dimensional case;
multidimensional extensions are a straightforward modification to
include in the CTU hydrodynamics scheme.  Our conserved variables are
\begin{equation}
  \Uc = \left ( \begin{array}{c}
           \rho \\
           \rho X_k \\
           \rho u \\
           \rho E \\
           \rho e \end{array}\right )
\end{equation}
where $\rho$ is the mass density, $u$ is velocity, $E$ is specific
total energy, $p$ is the pressure, and we carry nuclear species mass
fractions, $X_k$.  The specific total
energy relates to the specific internal energy, $e$, as $E = e + u^2/2$,
and we also separately evolve $e$, as part of a dual
energy formulation (see \citealt{bryan:1995,wdmergerI}).
 The mass fractions are constrained to sum to 1, $\sum_k X_k = 1$.
Defining the hydrodynamical fluxes:
\begin{equation}
  \Fb(\Uc) = \left ( \begin{array}{c}
         \rho u \\
         \rho X_k u \\
         \rho u^2 p \\
         (\rho E + p) u \\
         \rho e u \end{array}\right )
\end{equation}
We can write the system in conservative form for all state variables aside from $(\rho e)$ as:
\begin{equation}
  \ddt{\Uc} + \ddx{\Fb(\Uc)} = \Sc(\Uc)
\end{equation}
where for the special case of $(\rho e)$, we have an additional ``pdV'' term:
\begin{equation}
\ddt{(\rho e)} + \ddx{F(\rho e)} + p \ddx{u} = S(\rho e)
\end{equation}

@@

To close the system, we need the equation of state of the form:
\begin{equation}
p = p(\rho, X_k, e)
\end{equation}

@@

The CTU+PPM method for hydrodynamics is second-order accurate in space
and time.  We want to couple the reactive sources to the hydrodynamics
to be second-order in time as well.  As discussed above, nuclear
reaction sources are stiff, and need to be integrated using implicit
methods for stabilty.  Operator splitting (e.g., Strang) is
traditionally employed here, and is used as a benchmark for comparison
in this paper.  We'll discuss this traditional approach next before moving
on to our new time-coupling method.

In Strang splitting, we first integrate the system with reactions terms only (no advection)
over $\Delta t/2$, then integrate the advection terms only (no reactions) over $\Delta t$,
and finally integrate the reaction terms only over $\Delta t/2$.  This flow is illustrated
in Figure~\ref{fig:strang_flowchart}.

@@

@@

Since this method is based off of the CTU hydrodynamics scheme, it
benefits from the larger timestep that method can take (when done with
full corner coupling, the advective CFL condition is unity) as
compared to a method-of-lines approach (see \citealt{ppmunsplit}).  In
addition to the standard CFL timestep limiter for explicit
hydrodynamics, timestep limiters based on
the energy generation or abundance changes over a timestep, such as
those introduced in \cite{prometheus}, are often used.  An energy limiter takes the
form of either:
\begin{equation}
\label{eq:dt:nuce_I}
\Delta t \le f_e\, \min_{i} \left\{\frac{e^n_{i}}{|\dot{S}_{i}^n|}\right\} \\
\end{equation}
or
\begin{equation}
\label{eq:dt:nuce_II}
\Delta t \le f_e\, \min_{i} \left\{ \frac{e^n_i}{|\int \dot{S}_i(t) dt|} \right\}
\end{equation}
where $i$ is the zone index and $f_e$ is a parameter used to control
the allowed change (e.g., $f_e = 0.5$ will limit the step such that
reactions can only increase the internal energy by 50\%).  The two
formulations differ in that Eq.~\ref{eq:dt:nuce_I} uses the
instantaneous evaluation of the energy generation rate at the start of
the timestep while Eq.~\ref{eq:dt:nuce_II} uses the integral of the
energy generation rate over the entire step (or last Strang half).
Usually these limiters are used in a reactive manner---the next step will see
a smaller timestep based on what happened during the current step, but the
current step is still accepted, even if it violated these conditions.

\castro\ has the ability to reject a timestep if it detects a failure and retry
with smaller timesteps (subcycling to make up the original required timestep).
Among the conditions that can trigger this are density falling below zero during
advection, the ODE integration failing to converge in the implicit solve (due to
too many steps or the internal timestep falling below a minimum), or violation
of one of the timestep limiters during the step.  This means that the timestep
constraints are proactive instead of reactive.  The retry mechanism in \castro\
works with both the Strang and simplified-SDC integration scheme.  Retrying the
step is much more stringent that simply adjusting the next step, so we put a cap
on the number of retries allowed, aborting if we exceed the cap.

@@

Our simplified-SDC approach is based on the method described in \citet{SDC-old} and
a similar (but unpublished) implementation in the \maestroex\ simulation code \cite{maestroex}.
We again start by considering the update
in integral form:
\begin{equation}
\Uc^{n+1} = \Uc^n + \int_t^{t+\Delta t} \left [ \Advs{\Uc} + \Rb(\Uc) \right ] dt,
\end{equation}
In this approach, we 
approximate the advective term
piecewise constant in time, using the value at the midpoint in time,
to achieve second-order accuracy, giving us:
\begin{equation}
\label{eq:integral:simplesdc}
\Uc^{n+1} = \Uc^n + \int_t^{t+\Delta t} \left \{ \Adv{\Uc}^{n+1/2} + \Rb(\Uc) \right \} dt
\end{equation}
The \castro\ version of the simplified-SDC algorithm
differs from the previous versions due to the need to do some operations
on the conserved variable state and some on the primitive variable
state.   The overall flow of the simplified-SDC algorithm is shown
in Figure~\ref{fig:flowchart_II}.

@@

We use a modified version of the VODE~\citep{vode} integrator,
designed for stiff ODEs.  Our version has been ported C++, as a
header-only library enabling it to run on GPUs.  We have also added
checks to the timestep rejection logic that help prevent VODE from
wandering too far off the solution\footnote{This modified version of
VODE is available in our Microphysics repo:
\url{https://github.com/amrex-astro/Microphysics}.}  These have
been shown empirically to improve the performance of VODE.

@@

For Strang, we require that
an individual mass fraction not change too much over a step:
\begin{equation}
\label{eq:vode_species_reject}
\begin{array}{c}|X_k^{m+1}| < f |X_k^m| \\ |X_k^{m+1}| > \frac{1}{f} |X_k^m| \end{array} ~  \mbox{if}~ |X_k^m| > \eta \atolspec ~\mbox{and}~ |X_k^{m+1}| > \eta \atolspec 
\end{equation}
Here, $m$ is the current VODE solution and $m+1$ is the potential
new-time solution.  The parameter $f$ is the factor a mass fraction is
allowed to change per VODE step.  We use $f = 4$ for the results here.
The parameter $\eta$ allows us to only use this condition for species
with mass fractions above a threshold.  We use $\eta = 1$ by default,
but $\eta = f$ sometimes helps prevent VODE from worrying about
species that dip below the threshold during integration, since VODE
uses a norm over all of the integrated variables to compute the total
error.  If these conditions are violated, then VODE will reject the
timestep and retry with a smaller step.  We also require that the mass
fractions are contained in $[0, 1]$ to a tolerance of $10^{-2}$.
For SDC, we enforce these same constraints, but for
Eq.~\ref{eq:vode_species_reject}, we only use the change from $m$ to $m+1$ due
to reactions, by subtracting off the advective contribution over that substep.
We also require that:
\begin{align}
\rho^{m+1} &> 0 \\
(\rho e)^{m+1} &> 0
\end{align}

Finally, some of the rates in \citet{reaclib} increase by 100s of orders of
magnitude at low temperature, presumably because the interpolant fit was done at
higher temperatures.  Since VODE can take exploratory righthand side evaluations
outside of the nominal time range, it can encounter these poorly-behaved rates
and it can cause the integration to fail (by generating {\tt inf}s).  To prevent
this behavior, we set a lower temperature limit of the reaction rates, below
which we zero them out.  For the simulations here, we choose a cutoff of
$5\times 10^7~\mathrm{K}$.

For both Strang and simplified-SDC, we will use a numerical Jacobian that VODE
computes internally (following the algorithm in \citealt{lsode}).  For
completeness, we derive the analytic form of the Jacobian in Appendix
\ref{sec:app:jacobian} for the simplified-SDC method. VODE employs
Jacobian-caching, so the Jacobian does not need to be re-evaluated each time it
is used.

@@

\cite{castro_sdc} introduced a test problem for measuring convergence
of a reacting hydrodynamic algorithm.  This is based on the acoustic
pulse problem from \citep{mccorquodalecolella}, and was also used in
\citep{strang_rnaas}.  A periodic domain with a uniform entropy is
initialized with a pressure profile with a small perturbation.  This
drives a low amplitude acoustic wave radially outward from the center.
A simple reaction network with the triple-alpha and
$\isotm{C}{12}(\alpha,\gamma)\isotm{O}{16}$ rates releases energy.  By
running at different resolutions, with a fixed timestep that scales
with resolution, we can compute the convergence rate.  In order to
demonstrate second-order convergence, the problem must be smooth (so
the slope limiters don't kick in severely), which means that the
energy release cannot be so vigorous as to drive shocks.

@@

Tables \ref{table:react_converge_strang} and
\ref{table:react_converge_sdc} show the Strang and simplified-SDC
convergence respectively.  We see that both methods converge nearly
second order, and agree well with each other.  This demonstrates that both
methods are working as designed.

@@

We next model a double detonation Type Ia supernova.  In this model, a
detonation begins in the accreted helium layer on a sub-Chandrasekhar mass white
dwarf and that detonation can either drive a compression wave into the core
igniting a second detonation in the carbon / oxygen core, or burn through the
interface at the base of the accreted layer producing an inward propagating
carbon detonation (see, e.g., \citealt{fink:2007}).  Our goal here is not to
explore the feasibility of the model or understand whether a detonation
propagating through the He-C interface is physical, but rather to look at the
coupling of the hydrodynamics and reactions.  We want to drive vigorous burning
and directly compare Strang and simplified-SDC time integration on this problem.
For that reason, we explore only a single model and start off with a rather
large temperature perturbation in the He layer.

We use a network with 28 nuclei and 107 rates that captures He burning with
links for $(\alpha,p)(p,\gamma)$ reactions and the rates involving \isot{N}{14}
to bypass $\isotm{C}{12}(\alpha,\gamma)\isotm{O}{16}$ as described in
\cite{shenbildsten}.  The rates are taken from the ReacLib library
\citep{reaclib} and the network is written by \pynucastro~\citep{pynucastro}
directly in the C++ format that our code requires.  The details of the network
are given in Appendix \ref{sec:app:reactionnet}. This moderate-sized network
with many reaction pathways (forward and reverse) make it a good test for the
coupling between hydro and reactions.

@@

An initial perturbation is placed on the symmetry axis of the form:
\begin{equation}
  T = T_0 \left [ 1 + X(\isotm{He}{4}) f (1 + \tanh(2 - r1)) \right ]
\end{equation}
where
\begin{equation}
  r_1 = \left [ x^2 + (y - r_0)^2 \right ]^{1/2} / \lambda
\end{equation}
and
\begin{equation}
  r_0 = r_\mathrm{pert} + r_\mathrm{base}
\end{equation}
Here, $r_\mathrm{base}$ is the radius at which the helium layer begins
and $r_\mathrm{pert}$ is the distance above the base to put the perturbation.  We
choose $r_\mathrm{pert} = 100~\mathrm{km}$.  The temperature is pertubed above
the initial model value, denoted as $T_0$ here.  The amplitude of the perturbation
is $f = 12$ and the scale of the perturbation is $\lambda = 200~\mathrm{km}$.

We do a suite of runs with both Strang and the simplified-SDC integration, using
CFL numbers of 0.1, 0.2, and 0.4 and reaction network tolerances of $\atolspec =
\rtolspec = 10^{-5}$ and $\atolener = \rtolener = 10^{-65$.  We also run the CFL
= 0.2 case with a tighter tolerance, using  $\atolspec = 10^{-8}$.  For the step
rejection part of VODE, we use $f = 4$ and $\eta = 4$.

@@

\begin{deluxetable}{llll}
\tablecaption{\label{table:masses} Species masses.}
\tablehead{\colhead{simulation} & \colhead{$M_{\isotm{He}{4}}$} & \colhead{$M_{\isotm{Si}{28}}$} & \colhead{$M_{\isotm{Ni}{56}}$}}
\startdata
\multicolumn{4}{c}{Strang runs} \\
\hline
{\tt strang\_subch2\_cfl0.1}           & $1.059\times 10^{32}$  &  $3.804\times 10^{31}$  &  $1.092\times 10^{32}$ \\
{\tt strang\_subch2\_cfl0.2}           & $9.114\times 10^{31}$  &  $4.174\times 10^{31}$  &  $1.078\times 10^{32}$ \\
{\tt strang\_subch2\_cfl0.2\_tol1.e-8} & $7.222\times 10^{31}$  &  $8.623\times 10^{31}$  &  $4.041\times 10^{31}$ \\
{\tt strang\_subch2\_cfl0.4}           & $7.819\times 10^{31}$  &  $4.208\times 10^{31}$  &  $1.045\times 10^{32}$ \\
\hline
\multicolumn{4}{c}{SDC runs} \\
\hline
{\tt sdc\_subch2\_cfl0.1}           & $7.319\times 10^{31}$  &  $9.147\times 10^{31}$  &  $3.765\times 10^{31}$ \\
{\tt sdc\_subch2\_cfl0.2}           & $7.208\times 10^{31}$  &  $9.151\times 10^{31}$  &  $3.783\times 10^{31}$ \\
{\tt sdc\_subch2\_cfl0.2\_tol1.e-8} & $7.376\times 10^{31}$  &  $9.220\times 10^{31}$  &  $3.513\times 10^{31}$ \\
{\tt sdc\_subch2\_cfl0.4}           & $7.203\times 10^{31}$  &  $8.843\times 10^{31}$  &  $3.754\times 10^{31}$ \\
\hline
\enddata
\end{deluxetable}

@@

\section{Reaction Network}
\label{sec:app:reactionnet}
We build a reaction network that approximates an $\alpha$-chain,
including all $\isotm{He}{4}$, and all of the $\alpha$ nuclei from
$\isotm{C}{12}$ to $\isotm{Ni}{56}$.  We want to capture both the
$(\alpha,\gamma)$ and $(\alpha,p)(p,\gamma)$ links, so we include the
intermediate nuclei from the $(\alpha,p)$ rate instead of
approximating these links assuming proton equilibriation.  Following
\citet{shenbildsten}, we include the sequence
$\isotm{N}{14}(\alpha,\gamma)\isotm{F}{18}(\alpha, p)\isotm{Ne}{21}$,
which produce the protons that can allow for $\isotm{C}{12}(p,
\gamma)\isotm{N}{13}(\alpha, p)\isotm{O}{16}$ to compete with
$\isotm{C}{12}(\alpha,\gamma)\isotm{O}{16}$.  We add $\isotm{Na}{22}$
to link the $\isotm{Ne}{21}$ produced back to the $\alpha$-chain.  We
grab all of the ReacLib \citep{reaclib} reaction rates linking these
nuclei using \pynucastro~\citep{pynucastro}.  We are missing one rate
each for $\isotm{C}{12} + \isotm{C}{12}$, $\isotm{C}{12} +
\isotm{O}{16}$, and $\isotm{O}{16} + \isotm{O}{16}$.  These are
the sequences $\isotm{C}{12}(\isotm{C}{12}, \mathrm{n})\isotm{Mg}{23}(n, \gamma)\isotm{Mg}{24}$,
$\isotm{O}{16}(\isotm{C}{12}, \mathrm{n})\isotm{Si}{27}(n, \gamma)\isotm{Si}{28}$, and
$\isotm{O}{16}(\isotm{O}{16}, \mathrm{n})\isotm{S}{31}(n, \gamma)\isotm{S}{32}$,involving neutron
production followed by a capture back to one of our $\alpha$-chain
nuclei.  These are the only rates we approximate here, assuming
that the subsequent neutron capture is instantaneous.
